Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-09-19 paper/paper.tex
edits to pill section (no more fixmes!!!) Tested everything on bennet.

\begin{figure*}
  %\includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/image-plot}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/single-image-plot}\\
    \vspace{-2.5em}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/single-image-plot}
    \vspace{-3em}
  \end{center}
  \caption{Contour plot images of the concentration of each protein
    species in a natural pill-shaped bacterium at regular intervals in
    time (one second intervals), with darker regions indicating higher
    concentration. The upper plots shows results from the
    deterministic model and the lower shows results from the
    stochastic model.  The cells pictured are $4\micron$ in length,
    measured from end to end.  The order of frames is such that
    individual MinD proteins begin at the bottom of the plot (in the
    MinD:ATP state in the cytoplasm), and progress upward until they
    reach the MinE:MinD:ATP membrane-bound complex.  At that point,
    they will spontaneously dissociate into cytoplasmic MinE (the top
    row) and the starting state of cytoplasmic MinD:ADP.  The
    densities plotted are integrated along the axis orthogonal to the
    page.  The stochastic densities are broadened using the gaussian
    smearing approach of Zhang \emph{et
      al.}~\cite{zhang2007gaussian}}.
  \label{image-p}
\end{figure*}

Huang \emph{et al}~\cite{huang2003dynamic} has performed a linear
stability analysis on the model which shows an upper limit on a steady
state solution of a $2\micron$ wavelength.  Cells with dimensions longer
than this show spontaneous oscillatory behavoir in this
direction, while cells that are shorter relax into a motionless steady
state.  We have similarly performed a stability analysis in an
infinite slab with a thickness equal to that of our flattened cells
($0.25\micron$) and have found that an equivalent stability limit for
half wavelength is $2.13\micron$. As expected, when decreasing the
lengths and widths of our simulated flattened cells so that the
longest distance across the cell is less than this length, the cells
stop exhibiting any oscillatory behavior.  The deterministic model
relaxes into a motionless state and the stochastic model exhibits
random fluctuations without spatial oscillations.

@@

We begin with the naturally occuring pill cell shape.  We peice this
shape together as two hemispherical endcaps attached on either end of
a cylinder.  This shape follows the early simulations of Huang
\emph{et al.} but differs in that we have added the end caps for a
more natural shape, expecting similar results.

Figure~\ref{image-p} shows a series of color plots of protein cycle
stage concentration for both the deterministic and stochastic
models. The cells shown are $4\micron$ in length, measured from end to
end.  We have 'smeared out' the stochastic concentrations in a manner
meant to reproduce the images shown by diffraction limited
flourescence microscopy.  We do so using the two dimensional gaussian
approximation developed by Zhang \emph{et
  al.}~\cite{zhang2007gaussian}.  In this approximation we use a
numerical aperture value of 1.3, which is the same as used by Mannik
experimentally, and a wavelength of $650nm$.  Each frame is 2.5
seconds ahead of the last, and each image shows the concentration of a
given state of protein (of the five described in the reaction model)
summed over the coordinate orthogonal to the page.

Figure~\ref{image-p} begins about 300 seconds into the simulation and
shows one period of oscillation.  At $t=0$ there is a high
concentration of MinD:ATP that has accumulated on the membrane in the
lower section of the cell. An important aspect of Huang's model is
that the MinD is attracted to and sticks to the membrane nonlinearly:
as it accumulates there it begins to 'recruit' other MinD that is
diffusing in the cytoplasm nearby, causing peaks in concentration to
build up on the walls.  Meanwhile, the MinE is creeping downward as it
reacts with the membrane-bound MinD, forming the MinD:MinE:ATP
complex, breaks it apart, and then diffuses downward a bit more before
it again reacts with membrane-bound MinD.  This process can seen by
the famous MinE 'rings' shown in the deterministic model plots, from 0
seconds to 4 seconds.  The appearence of these rings in the
deterministic model along side their less convincing appearence in the
stochastic model highlights an advantage of the deterministic model:
the idealization of deterministic data allows one to see macroscopic
patterns that might otherwise be missed.  In the stochastic model the
MinE still exhibits higher concentrations in the same regions
throughout the process, but what would ideally be a 'ring' pattern is
instead an assymetric collection of maxima.

During the formation of these rings, MinE that is diffusing in the
cytoplasm in the upper portion of the cell will naturally diffuse
downward, where there is membrane-bound MinD to react with, causing a
depletion of MinE in the upper portion of the cell.  As the MinE ring
converges upon the lower end of the cell, MinD that has been released
is able to diffuse upward, past the ring, while still in its MinD:ADP
state.  It then undergoes nucleotide exchange, resulting in MinD:ATP,
which is ready to accumulate on the walls in the upper portion of the
cell, where the MinE has been depleted.  This can be seen in seconds
10 through 20 in both models, and the subsequent MinE ring
formation and movement upward (beginning the same process in the
opposite direction) can be seen in seconds 15 through 23.

@@

Figure~\ref{corr-pill} shows a plot of the time correlation between
the total MinD found in the two opposite polar regions.  This
normalized value is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the outer sections of a cell that has been divided
into thirds along its long axis.  These plots heolp us in studying the
periods and the regularity and stability of oscillations.  We have
fitted them to a simple exponential decay model (akin to that of
monochromatic light), given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of over
7 periods and in the case of the determinisitic model oscillate
indefinately with complete coherence.

\fixme{Do we want anything from the following paragrph?}While the
stochastic results show periods that seem to have a high degree of
regularity as well as similarity to the deterministic results, they do
differ in the precise locations of the maxima.  While the
deterministic model shows maxima that are every time concentrated in
the center of one of the cell ends, with a symmetric smear of protein
going out laterally to either side, the stochastic model shows maxima
that, while always located somehwere in the polar region of the cell,
occur at different positions within this region.  This is reflected in
Figure~\ref{image-p}, which shows periods that are the same and
qualitative behavoir that is approximately the same, but also that
there is variation in the details of the stochastic model maxima
location.  \fixme{I think we should discuss stochastic picture
  earlier.}

>>>>

\begin{figure*}
  %\includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/image-plot}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/single-image-plot}\\
    \vspace{-2.5em}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/single-image-plot}
    \vspace{-3em}
  \end{center}
  \caption{Images of the concentration of each protein species in a
    natural pill-shaped bacterium at one-second time intervals. The
    upper plots shows results from the deterministic model and the
    lower shows results from the stochastic model.  The order of
    frames is such that individual MinD proteins begin at the bottom
    of the plot (in the MinD:ATP state in the cytoplasm), and progress
    upward until they reach the MinE:MinD:ATP membrane-bound complex.
    At that point, they will spontaneously dissociate into cytoplasmic
    MinE (the top row) and the starting state of cytoplasmic
    MinD:ADP.}.
  \label{image-p}
\end{figure*}

Huang \emph{et al.}~\cite{huang2003dynamic} has performed a linear
stability analysis on the model which shows an upper limit on a steady
state solution of a $2\micron$ wavelength.  Cells with dimensions longer
than this show spontaneous oscillatory behavoir in this
direction, while cells that are shorter relax into a motionless steady
state.  We have similarly performed a stability analysis in an
infinite slab with a thickness equal to that of our flattened cells
($0.25\micron$) and have found that an equivalent stability limit for
half wavelength is $2.13\micron$. As expected, when decreasing the
lengths and widths of our simulated flattened cells so that the
longest distance across the cell is less than this length, the cells
stop exhibiting any oscillatory behavior.  The deterministic model
relaxes into a motionless state and the stochastic model exhibits
random fluctuations without spatial oscillations.

@@

We begin with the naturally occuring pill cell shape.  We piece this
shape together as a cylinder with hemispherical endcaps.  This shape
follows the early simulations of Huang \emph{et al.} but differs in
that we have added the end caps for a more natural shape, expecting
similar results.

Figure~\ref{image-p} shows a series of color plots of the density of
proteins at each stage of the reaction cycle. The cells shown are
$4\micron$ in length, measured from end to end.  We have `smeared out'
the stochastic concentrations in a manner meant to reproduce the
images shown by diffraction limited flourescence microscopy.  We do so
using the two dimensional gaussian approximation developed by Zhang
\emph{et al.}~\cite{zhang2007gaussian}.  In this approximation we use
a numerical aperture value of 1.3, which is the same as used by Mannik
experimentally, and a wavelength of $650nm$.  Each frame is 2.5
seconds ahead of the last, and each image shows the concentration of a
given state of protein (of the five described in the reaction model)
summed over the coordinate orthogonal to the page.

Figure~\ref{image-p} begins about 300 seconds into the simulation and
shows one period of oscillation.  At $t=0$ there is a high
concentration of MinD:ATP that has accumulated on the membrane at the
bottom of the cell. An important aspect of Huang's model is that the
MinD is attracted to and sticks to the membrane nonlinearly: as it
accumulates there it begins to `recruit' other MinD that is diffusing
in the cytoplasm nearby, causing peaks in concentration to build up on
the walls.  Meanwhile, the MinE is creeping downward as is bound to
the membrane-bound MinD, forming the MinD:MinE:ATP complex, then
breaks it apart and diffuses downward a bit more before it again
reacts with membrane-bound MinD.  This process can seen in the form of
the well-known ``MinE rings'' (actually, MinE bound to MinD on the
membrane).  These rings are visible in the deterministic model plots
as a green band on the walls from 0 seconds to 4 seconds (and later
from 15 to 23 seconds).  The appearence of these rings in the
deterministic model along side their less obvious appearence in the
stochastic model highlights an advantage of the deterministic model:
the idealization of deterministic data allows one to see patterns in
the averaged behavior that might otherwise be missed.  In the
stochastic model the MinE still exhibits higher concentrations in the
same regions throughout the process, but what would ideally be a
``ring'' pattern is instead an asymmetric collection of maxima, which
would become a ring after phase-locked averaging.

During the formation of these rings, MinE that is diffusing in the
cytoplasm in the upper portion of the cell will naturally diffuse
downward, where there is membrane-bound MinD to react with, leading to a
depletion of MinE in the upper portion of the cell.  As the MinE ring
converges upon the lower end of the cell, MinD that has been released
is able to diffuse upward, past the ring, while still in its MinD:ADP
state.  After it undergoes nucleotide exchange, resulting in MinD:ATP,
it is ready to accumulate on the walls in at the top of the
cell, where the MinE has been depleted.  This can be seen in seconds
10 through 20 in both models, followed by the subsequent MinE ring
formation and movement upward (beginning the same process in the
opposite direction) can be seen in seconds 15 through 23.

@@

It is clear from Fig.~\ref{image-p} that the stochastic simulation
results in a protein distribution that is spatially regular than the
deterministic model predicts.  In fact, when viewed as a movie (see
supplementary material), the stochastic model exhibits a ``starry
night'' effect, as recruitment leads to clusters of MinD forming on
the wall and then subsequently dissipating.  This naturally leads to
irregularity in the temporal periodicity of the system, which we
quantify by examining the temporal correlation function of the total
MinD found in two opposite polar regions.
%
This correlation function, which is displayed in Fig.~\ref{corr-pill},
is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom sections of a cell that has been
divided into thirds along its long axis.  These plots help us in
studying the periods and the regularity and stability of oscillations.
These curves are well fit with a simple exponential decoherence model,
given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 7 periods and in the case of the determinisitic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.
